---
title: "Anxiety Z-score linear model"
author: "Elena Eyre, PhD"
date: '2022-02-01'
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: united
  word_document:
    toc: yes
  pdf_document:
    toc: yes
#urlcolor: tomato
linkcolor: steelblue2
fontsize: 8 pt
---

# 1. Loadings

## 1.1 Load packages

I uploaded packages of necessary functions to perform the calculations.

```{r, setup}
knitr::opts_chunk$set(warning=FALSE)
package_to_load <- c("readxl", "dplyr", "ggplot2", "tidyr", "dplyr","RColorBrewer")
for (package in package_to_load) {
  require(package, character.only = T); packageVersion(package)
}
extra_to_load <- c("knitr","stringr", "stringi", "ggrepel","ggpubr", "ggbreak","reshape2","lme4","multcomp","car","lsmeans")
for (package in extra_to_load) {
  require(package, character.only = T); packageVersion(package)
}
#rm(package_to_load, extra_to_load)
```

## 1.2 Load data

```{r}
df_original <- read_excel("220128_Analysis_adapted_Elena.xlsx")
anxiety_raw_dlb <- df_original[df_original$Experiment == "Dark Light Box",]
anxiety_raw_epm <- df_original[df_original$Experiment == "Elevate Pluse Maze",]
```

# 2. Previous to Z-score

Before the Z-score, we need to calculate control means and standard deviation.

For the Dark-light box experiment I make sure to select the center type:
```{r}
male_c_mean_dlb <- mean(anxiety_raw_dlb$Value[anxiety_raw_dlb$Sex == "M" & anxiety_raw_dlb$Diet == "CTRL"]) 
male_c_sd_dlb <- sd(anxiety_raw_dlb$Value[anxiety_raw_dlb$Sex == "M" & anxiety_raw_dlb$Diet == "CTRL"])
female_c_mean_dlb <- mean(anxiety_raw_dlb$Value[anxiety_raw_dlb$Sex == "F" & anxiety_raw_dlb$Diet == "CTRL"])
female_c_sd_dlb <- sd(anxiety_raw_dlb$Value[anxiety_raw_dlb$Sex == "F" & anxiety_raw_dlb$Diet == "CTRL"])
```

For the Elevated Plus Maze adulthood experiment:
```{r}
male_c_mean_epm <- mean(anxiety_raw_epm$Value[anxiety_raw_epm$Sex == "M" & anxiety_raw_epm$Diet == "CTRL"]) 
male_c_sd_epm <- sd(anxiety_raw_epm$Value[anxiety_raw_epm$Sex == "M" & anxiety_raw_epm$Diet == "CTRL"])
female_c_mean_epm <- mean(anxiety_raw_epm$Value[anxiety_raw_epm$Sex == "F" & anxiety_raw_epm$Diet == "CTRL"])
female_c_sd_epm <- sd(anxiety_raw_epm$Value[anxiety_raw_epm$Sex == "F" & anxiety_raw_epm$Diet == "CTRL"])
```

# 3. Calculate the Z-scores for each subset and experiment

To speed up the process, I created a function to calculate the z-score: zscore_gender(). It will return a vector of values corresponding to the z-scores associated to the subset we selected.
```{r}
zscore_gender <- function(data_frame,gender, mean_control,sd_control){
  d <- as.data.frame(data_frame[data_frame$Sex == gender & data_frame$Diet == "Emul",])
  d$z_score_male_of <-sapply(d$Value, function(x){
  (x - mean_control)/sd_control
  })
  return(d)
  }
```

For the Dark-light box experiment:
```{r}
z_score_male_dlb <-  zscore_gender(anxiety_raw_dlb, "M", male_c_mean_dlb, male_c_sd_dlb)
z_score_female_dlb <-  zscore_gender(anxiety_raw_dlb, "F", female_c_mean_dlb, female_c_sd_dlb)
```

For the Elevated Plus Maze adulthood (12-13 weeks) experiment:
```{r}
z_score_male_epm <-  zscore_gender(anxiety_raw_epm, "M", male_c_mean_epm, male_c_sd_epm)
z_score_female_epm <-  zscore_gender(anxiety_raw_epm, "F", female_c_mean_epm, female_c_sd_epm)
```

# 4. Visualize the Z-scores for each subset and experiment

In order to siplify the results, we join all the results into a new table, save it in a .txt file, and split it in order to define the data we want to show in a plot:
```{r}
df1 <- data.frame(rbind(z_score_male_dlb,z_score_female_dlb,z_score_male_epm,z_score_female_epm))
names(df1)[8] <- "Z_score"
kable(df1[1:4,],format = "simple",caption = "Reordered data table (with mothers and pups)") #%>% kable_styling(full_width = F, position = "left")%>%column_spec(1, bold = T)
write.table(df1, "Values_gender_Zscores_all_anxiety_experiments_withMothers_and_diets.txt", sep = "|", dec = ".", row.names = F)

df2_dlb <- df1[df1$Experiment == "Dark Light Box", ]
df2_epm <- df1[df1$Experiment == "Elevate Pluse Maze", ]
```

Plots:
```{r, fig.width=12, fig.height=4}
set.seed(123) # Due to I apply a statistics analysis to extract the p-value from a t-test, I make sure to not bias it by adding this se.seed( ) function.
g2 <- ggplot(df2_dlb, aes(x=as.factor(Sex), y=Z_score)) + geom_boxplot(aes(x=as.factor(Sex),fill = as.factor(Sex))) + xlab("Sex") + ylab("Z-score") + ggtitle("Dark-light box experiment") + theme_classic()+ labs(fill= "Sex")+ stat_compare_means(method = "t.test")
g3 <- ggplot(df2_epm, aes(x=as.factor(Sex), y=Z_score)) + geom_boxplot(aes(x=as.factor(Sex),fill = as.factor(Sex))) + xlab("Sex") + ylab("Z-score") + ggtitle("Elevated Plus Maze experiment") + theme_classic()+ labs(fill= "Sex")+ stat_compare_means(method = "t.test")
ggarrange(g2, g3, ncol = 3, nrow = 1)
pdf("Boxplots_gender_Zscores_anxiety_experiments_individually_MP.pdf", width = 5, height = 4)
g2
g3
dev.off()
pdf("Boxplots_gender_Zscores_all_anxiety_experiments_MP.pdf", width = 11, height = 4)
ggarrange(g2, g3, ncol = 3, nrow = 1)
dev.off()
##Before calculating the z-score:
g5 <- ggplot(anxiety_raw_dlb, aes(x=as.factor(Diet), y=Value)) + geom_boxplot(aes(x=as.factor(Diet),fill = as.factor(Sex))) + xlab("Diet") + ylab("Time spent in light (%)") + ggtitle("Dark-light box adulthood experiment") + theme_classic()+ labs(fill= "Sex")#+ stat_compare_means(method = "t.test")
g6 <- ggplot(anxiety_raw_epm, aes(x=as.factor(Diet), y=Value)) + geom_boxplot(aes(x=as.factor(Diet),fill = as.factor(Sex))) + xlab("Diet") + ylab("Time in zone (%)") + ggtitle("Elevated Plus Maze experiment") + theme_classic()+ labs(fill= "Sex")#+ stat_compare_means(method = "t.test")
pdf("Boxplots_gender_raw_data_anxiety_experiments_MP.pdf", width = 11, height = 4)
ggarrange(g5, g6, ncol = 3, nrow = 1)
dev.off()
ggarrange(g5, g6, ncol = 3, nrow = 1)
```

# 5. Anxiety Z-score for only Females vs Males

I explore the outcome if we take the Chow group as the corresponding control group, and we would like to compare only Gender, I develop another z-score formula:

Then, we follow the line as Guilloux et al did in their paper (doi:10.1016/j.jneumeth.2011.01.019), and we callculated individual anxiety scores by averaging z-score values across tests (leveraging the potential biases nduced by a single test). As they did, an emotionally z-score was calculated for each animal based on 3 different tests:

$Anxiety~Score = (Z_DLB + Z_EPM) / Number~of~tests$

```{r, fig.width=5, fig.height=4}
df3m <- NULL
df3f <- NULL
df3m$Anx_Zscore <- (df1$Z_score[df1$Experiment == "Dark Light Box" & df1$Sex == "M"] + df1$Z_score[df1$Experiment == "Elevate Pluse Maze" & df1$Sex == "M"])/2
df3f$Anx_Zscore <- ( df1$Z_score[df1$Experiment == "Dark Light Box" & df1$Sex == "F"] + df1$Z_score[df1$Experiment == "Elevate Pluse Maze" & df1$Sex == "F"])/2
df3m$Sex <- "M"
df3f$Sex <- "F"
df3 <- as.data.frame(rbind(as.data.frame(df3m),as.data.frame(df3f)))
#df3 <- reshape2::melt(df3,variable.name = "Subsets",value.name = "Anx_Zscore")
write.table(df3, "Zscores_all_anxiety_experiments.txt", sep = "|", dec = ".", row.names = F)
g7 <- ggplot(df3, aes(x=as.factor(Sex), y=Anx_Zscore)) + geom_boxplot(aes(x=as.factor(Sex),fill = as.factor(Sex))) + xlab("Sex") + ylab("Anxiety Score\n(Z-score)") + ggtitle("Anxiety score") + theme_classic()+ labs(fill= "Sex")+ stat_compare_means(method = "t.test")
g7
pdf("Boxplots_gender_anxiety_final_score_experiments.pdf", width = 5, height = 4)
g7
dev.off()
```

# 6. Anxiety Z-score for Emul vs CT (and Gender)

Since the authors of the paper take the males group and the corresponding control group, and we would like to compare only diets, I develop another z-score formula:

```{r}
zscore <- function(data_frame,mean_control,sd_control){
  d <- as.data.frame(data_frame)
  d$z_score_male_of <-sapply(d$Value, function(x){
  (x - mean_control)/sd_control
  })
  }
```

Here I apply the formula using the males group and the corresponding control group, to each experiment:

```{r}
anxiety_raw_dlb$zscores <- zscore(anxiety_raw_dlb,male_c_mean_dlb,male_c_sd_dlb)
anxiety_raw_epm$zscores <- zscore(anxiety_raw_epm,male_c_mean_epm,male_c_sd_epm)

df5 <- rbind(anxiety_raw_dlb,anxiety_raw_epm)
df5 <- df5[df5$Sex != "F",]
df6 <- data.frame(DLB = df5[df5$Experiment =="Dark Light Box", ],EPM = df5[df5$Experiment =="Elevate Pluse Maze", ])
df6$Anx_zscore <- (df6$DLB.zscores + df6$EPM.zscores)/3
df6 <- df6[,c(2,3,6,7,17)] 
colnames(df6) <- c("Sex","Treatment","Mother", "Diet","Anx_zscore")

ggplot(df6, aes(x=as.factor(Diet), y=Anx_zscore)) + geom_boxplot(mapping= aes(x=as.factor(Diet), y=Anx_zscore,fill=as.factor(Diet)))+ xlab("Diet") + ylab("Anxiety Score\n(Z-score)")+labs(fill="Diet") + ggtitle("Anxiety score") +geom_point(pch = 21, position = position_jitterdodge(), aes(x=as.factor(Diet),fill = as.factor(Diet)))+ theme_classic()+ stat_compare_means(aes(x=as.factor(Diet), y=Anx_zscore),method = "t.test")

write.table(df6, "Anxiety_Zscores_only_males.txt", sep = "|", dec = ".", row.names = F)
```

Then, I apply the formula using the both, males and females group and the corresponding control group, to each experiment: the final outcome suggest there is no significant changes.

$Value = (1 | Mother) + Sex + Diet + Sex*Diet$

```{r, fig.width=9, fig.height=4}
anxiety_raw_dlb$zscores <- ifelse(anxiety_raw_dlb$Sex == "M", zscore(anxiety_raw_dlb,male_c_mean_dlb,male_c_sd_dlb), zscore(anxiety_raw_dlb,female_c_mean_dlb,female_c_sd_dlb))
anxiety_raw_epm$zscores <- ifelse(anxiety_raw_epm$Sex == "M", zscore(anxiety_raw_epm,male_c_mean_epm,male_c_sd_epm), zscore(anxiety_raw_epm,female_c_mean_epm,female_c_sd_epm))

df5 <- rbind(anxiety_raw_dlb,anxiety_raw_epm)
df6 <- data.frame(DLB = df5[df5$Experiment =="Dark Light Box", ],EPM = df5[df5$Experiment =="Elevate Pluse Maze", ])
df6$Anx_zscore <- (df6$DLB.zscores + df6$EPM.zscores)/3
df6 <- df6[,c(2,3,6,7,17)] 
colnames(df6) <- c("Sex","Treatment","Mother", "Diet","Anx_zscore")
write.table(df6, "Anxiety_Zscores_both_malesAndFemales.txt", sep = "|", dec = ".", row.names = F)

ggplot(df6, aes(x=as.factor(Sex), y=Anx_zscore)) + geom_boxplot(mapping= aes(x=as.factor(Sex), y=Anx_zscore,fill=as.factor(Diet)))+scale_y_continuous() + xlab("Sex") + ylab("Anxiety Score\n(Z-score)")+labs(fill="Diet") + ggtitle("Anxiety score") +geom_point(pch = 21, position = position_jitterdodge(), aes(x=as.factor(Sex),fill = as.factor(Diet)))+ theme_classic()+stat_compare_means(aes(x=as.factor(Sex), y=Anx_zscore,fill = as.factor(Diet)),method = "t.test")

fit_mf3 <- lmer(Anx_zscore ~ (1 | Mother) + Sex +  Diet + Sex*Diet, data=df6) 
Anova(fit_mf3, test ="Chisq", type="III", error.estimate = "deviance")
print(summary(fit_mf3))
tuk <- glht(fit_mf3, limfct = mcp(Diet = "Tukey")) # To obtain the details of the single variable
print(summary(tuk))
tuk <- glht(fit_mf3, limfct = mcp(Sex = "Tukey")) # To obtain the details of the single variable
print(summary(tuk))
plot(confint(tuk)) 
kable(as.data.frame(lsmeans(fit_mf3, pairwise~Sex*Diet, adjust="tukey")$contrasts)[,c(1,6)], caption = "Pairwise comparisons")
```

